A Certified Interpreter for the List Machine Benchmark

### 2020-06-08 papers/SBLP2020/paper.tex
Some improvements on text. Put Samuels coding of list benchmark rules

\usepackage{booktabs} % For formal tables
\usepackage[utf8x]{inputenc}
\usepackage{ucs}
\usepackage{graphicx}
\usepackage{amsmath,amsthm}
\usepackage{amssymb}
\usepackage{url}
\usepackage{stmaryrd}
\usepackage{ifpdf}
\ifpdf
\usepackage{hyperref}
\fi
\usepackage{float}
\usepackage{proof}

@@

\begin{abstract}
This is the abstract...
\end{abstract}

@@

\section{An Overview of Agda}\label{sec:agda}

Agda is a dependently-typed functional programming language based on
Martin-L\"of intuitionistic type theory~\cite{Lof98}.  Function types
and an infinite hierarchy of types of types, \ensuremath{\D{Set}\;\V{l}}, where \ensuremath{\V{l}} is a
natural number, are built-in. Everything else is a user-defined
type. The type \ensuremath{\D{Set}}, also known as \ensuremath{\D{Set}_{\D{0}}}, is the type of all
``small'' types, such as \ensuremath{\D{Bool}}, \ensuremath{\D{String}} and \ensuremath{\D{List}\;\D{Bool}}.  The type
\ensuremath{\D{Set}_{\D{1}}} is the type of \ensuremath{\D{Set}} and ``others like it'', such as \ensuremath{\D{Set}\;\to \;\D{Bool}}, \ensuremath{\D{String}\;\to \;\D{Set}}, and \ensuremath{\D{Set}\;\to \;\D{Set}}. We have that \ensuremath{\D{Set}\;\V{l}} is an
element of the type \ensuremath{\D{Set}\;(\V{l+1})}, for every $l \geq 0$. This
stratification of types is used to keep Agda consistent as a logical
theory~\cite{Sorensen2006}.

@@

proof of \ensuremath{\V{P}} for the element at position \ensuremath{\V{v}\;\Con{::}\;\D{Fin}\;\V{n}} in a \ensuremath{\D{Vec}}:

@@

>>>>

\usepackage{booktabs} % For formal tables
\usepackage[utf8x]{inputenc}
\usepackage{ucs}
\usepackage{graphicx}
\usepackage{amsmath,amsthm}
\usepackage{amssymb}
\usepackage{url}
\usepackage{stmaryrd}
\usepackage{ifpdf}
\usepackage{mathtools}
\usepackage{semantic}

\ifpdf
\usepackage{hyperref}
\fi
\usepackage{float}
\usepackage{proof}

@@

\begin{abstract}
Formal models are an important tool in the programming languages research
community. However, such models are full of intricacies and, due to that,
they are subject to subtle errors. Such failures motivated the usage of
tools to ensure the correctness of these formalisms. One way to eliminate
such errors is to encode models in a dependently-typed language in order
to ensure ``correctness-by-construction''.

In this work, we use follow this idea to build the verified interpreter
for the list machine benchmark in the Agda programming language.
\end{abstract}

@@

The development of new programming language design, linguistic construct
or type system involves its careful formalization in order to express
its core ideas in a concise way. However, such models have many details
and complexities which hinders its correctness assurances.
Because of such problems, the programming languages research community
started to use tools, like proof assistants~\cite{Stump16,Chlipala13},
and benchmark problems to validate them and stress its suitability for
such tasks~\cite{Aydemir05,Pientka18,Appel07}.

The rest of this paper is organized as follows. Section~\ref{sec:background}
presents a brief introduction to Agda and reviews the list machine benchmark.
In Section~\ref{sec:typing}, we
describe the type system for PEGs and its relation with the original
well-formedness predicate proposed by Ford~\cite{Ford2004}.
Section~\ref{sec:interpreter} describes the intrinsically typed syntax for PEGs
and its interpreter. Limitations of our approach are discussed
in Section~\ref{sec:discussion}. Related work is discussed in
Section~\ref{sec:related}, and Section~\ref{sec:conclusion} concludes.

All the source code in this article has been formalized in Agda
version 2.6.0 using Standard Library 1.0. All source code produced,
including the \LaTeX~ source of this
article, are available on-line~\cite{peg-rep}.

\section{Background}\label{sec:background}

\paragraph{An Overview of Agda}
Agda is a dependently-typed functional programming language based on
Martin-L\"of intuitionistic type theory~\cite{Lof98}.  Function types
and an infinite hierarchy of types of types, \ensuremath{\D{Set}\;\V{l}}, where \ensuremath{\V{l}} is a
natural number, are built-in. Everything else is a user-defined
type. The type \ensuremath{\D{Set}}, also known as \ensuremath{\D{Set}_{\D{0}}}, is the type of all
``small'' types, such as \ensuremath{\D{Bool}}, \ensuremath{\D{String}} and \ensuremath{\D{List}\;\D{Bool}}.  The type
\ensuremath{\D{Set}_{\D{1}}} is the type of \ensuremath{\D{Set}} and ``others like it'', such as \ensuremath{\D{Set}\;\to \;\D{Bool}}, \ensuremath{\D{String}\;\to \;\D{Set}}, and \ensuremath{\D{Set}\;\to \;\D{Set}}. We have that \ensuremath{\D{Set}\;\V{l}} is an
element of the type \ensuremath{\D{Set}\;(\V{l+1})}, for every $l \geq 0$. This
stratification of types is used to keep Agda consistent as a logical
theory~\cite{Sorensen2006}.

@@

proof of \ensuremath{\V{P}} for the element at position \ensuremath{\V{v}\;\mathbin{:}\;\D{Fin}\;\V{n}} in a \ensuremath{\D{Vec}}:

@@

\paragraph{The List Machine Benchmark}

\begin{equation}
\inference{}
          {(r, (\iota_1;\iota_2);\iota_3) \xmapsto{p} (r, \iota_1;(\iota_2;\iota_3))}[step-seq]
\end{equation}

\begin{equation}
\inference{r(v) = cons(a_0,a_1)~~~r[v':=a_0]=r'}
          {(r, (\textbf{fetch-field}~v~0~v';\iota)) \xmapsto{p} (r', \iota)}[step-fetch-field-0]
\end{equation}

\begin{equation}
\inference{r(v) = cons(a_0,a_1)~~~r[v':=a_1]=r'}
          {(r, (\textbf{fetch-field}~v~1~v';\iota)) \xmapsto{p} (r', \iota)}[step-fetch-field-1]
\end{equation}

\begin{equation}
\inference{r(v_0) = a_0~~~r(v_1)=a_1~~~r[v':=cons(a_0,a_1)]=r'}
          {(r, (\textbf{cons}~v_0~v_1~v';\iota)) \xmapsto{p} (r', \iota)}[step-cons]
\end{equation}

\begin{equation}
\inference{r(v) = cons(a_0,a_1)}
          {(r, (\textbf{branch-if-nil}~v~l;\iota)) \xmapsto{p} (r, \iota)}[step-branch-not-taken]
\end{equation}

\begin{equation}
\inference{r(v) = nil~~~p(l)=\iota'}
          {(r, (\textbf{branch-if-nil}~v~l;\iota)) \xmapsto{p} (r, \iota')}[step-branch-taken]
\end{equation}

\begin{equation}
\inference{p(l)=\iota'}
          {(r, \textbf{jump}~l) \xmapsto{p} (r, \iota')}[step-jump]
\end{equation}

\begin{equation}
\inference{(r, \iota) \xmapsto{p} (r', \iota')~~~(p, r', \iota')}
          {(p, r, \iota) \Downarrow}[run-step]
\end{equation}

\begin{equation}
\inference{}
          {(p, r, \textbf{halt}) \Downarrow}[run-halt]
\end{equation}

\begin{equation}
\inference{\{\}[\textbf{v}_0:=\text{nil}]=r~~~p(\textbf{L}_0)=\iota~~~(p,r,\iota)\Downarrow}
          {p \Downarrow}[run-prog]
\end{equation}

%\subsection{Typing}

%\subsubsection{Subtyping}

\begin{equation}
\inference{}
          {\tau \subset \tau}[subtype-refl]
\end{equation}

\begin{equation}
\inference{}
          {\text{nil} \subset \text{list}~\tau}[subtype-nil]
\end{equation}

\begin{equation}
\inference{\tau \subset \tau'}
          {\text{list}~\tau \subset \text{list}~\tau'}[subtype-list]
\end{equation}

\begin{equation}
\inference{\tau \subset \tau'}
          {\text{listcons}~\tau \subset \text{list}~\tau'}[subtype-listcons]
\end{equation}

\begin{equation}
\inference{\tau \subset \tau'}
          {\text{listcons}~\tau \subset \text{listcons}~\tau'}[subtype-listmixed]
\end{equation}

%\subsubsection{Instruction typing}

\begin{equation}
\inference{\Pi \vdash_{\text{instr}} \Gamma\{ \iota_1 \}\Gamma'~~~\Pi \vdash_{\text{instr}} \Gamma'\{ \iota_2 \}\Gamma''}
          {\Pi \vdash_{\text{instr}} \Gamma\{ \iota_1 ; \iota_2 \}\Gamma''}[check-instr-seq]
\end{equation}

\begin{equation}
\inference{\Gamma(v) = \text{list}~\tau~~~\Pi(l) = \Gamma_1~~~\Gamma[v:=\text{nil}]=\Gamma'~~~\Gamma' \subset \Gamma_1}
          {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{branch-if-nil}~v~l \}(v: \text{listcons}~\tau,~\Gamma')}[check-instr-branch-list]
\end{equation}

\begin{equation}
\inference{\Gamma(v) = \text{listcons}~\tau~~~\Pi(l) = \Gamma_1~~~\Gamma[v:=\text{nil}]=\Gamma'~~~\Gamma' \subset \Gamma_1}
          {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{branch-if-nil}~v~l \}\Gamma}[check-instr-branch-listcons]
\end{equation}

\begin{equation}
\inference{\Gamma(v) = \text{nil}~~~\Pi(l) = \Gamma_1~~~\Gamma \subset \Gamma_1}
          {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{branch-if-nil}~v~l \}\Gamma}[check-instr-branch-nil]
\end{equation}

\begin{equation}
\inference{\Gamma(v) = \text{listcons}~\tau~~~\Gamma[v':=\tau]=\Gamma'}
          {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{fetch-field}~v~0~v' \}\Gamma'}[check-instr-fetch-0]
\end{equation}

\begin{equation}
\inference{\Gamma(v) = \text{listcons}~\tau~~~\Gamma[v':=\text{list}~\tau]=\Gamma'}
          {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{fetch-field}~v~1~v' \}\Gamma'}[check-instr-fetch-1]
\end{equation}

\begin{equation}
\inference{\Gamma(v_0) = \tau_0~~~~~~~~~\Gamma(v_1) = \tau_1 \\ (\text{list}~\tau_0) \sqcap \tau_1=\text{list}~\tau~~~\Gamma[v:=\text{listcons}~\tau]=\Gamma'}
          {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{cons}~v_0~v_1~v \}\Gamma'}[check-instr-cons]
\end{equation}

%\subsubsection{Block typings}

\begin{equation}
\inference{}
          {\Pi;\Gamma\vdash_{\text{block}} \textbf{halt}}[check-block-halt]
\end{equation}

\begin{equation}
\inference{\Pi\vdash_{\text{instr}} \Gamma\{\iota_1\}\Gamma'~~~\Pi;\Gamma'\vdash_{\text{block}} \iota_2}
          {\Pi;\Gamma\vdash_{\text{block}} \iota_1;\iota_2}[check-block-seq]
\end{equation}

\begin{equation}
\inference{\Pi(l)=\Gamma_1~~~\Gamma \subset \Gamma_1}
          {\Pi;\Gamma\vdash_{\text{block}} \textbf{jump}~l}[check-block-jump]
\end{equation}

%\subsubsection{Program typings}

\begin{equation}
\inference{\Pi(l)=\Gamma~~~\Pi;\Gamma\vdash_{\text{block}} \iota~~~\Pi\vdash_{\text{blocks}} p}
          {\Pi\vdash_{\text{blocks}} l: \iota;~p}[check-blocks-label]
\end{equation}

\begin{equation}
\inference{}
          {\Pi\vdash_{\text{blocks}} \textbf{end}}[check-blocks-empty]
\end{equation}
